Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-06 treestats_paper.tex
Attempt at segue sentence b/w paragraphs.
Remove superlatives.

\krt{Why? I think this is an odd way of putting it.  The challenge back then was
    genotyping.  Once that was done, summarizing the relationships was done on a 
    calculator b/c the data sets were so small.}
It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.

@@

The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change because of the results of recombination in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
\citep{kelleher2016efficient} used this redundancy
to store genome sequence and the associated genealogical relationships extremely compactly,
as well as to compute several population genetics statistics highly efficiently.
This paper generalizes the strategy used by \citep{kelleher2016efficient} for statistics computation
to a much broader class of statistics.

>>>>

\krt{Why? I think this is an odd way of putting it.  The challenge back then was
    genotyping.  Once that was done, summarizing the relationships was done on a 
    calculator b/c the data sets were so small.}
It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
To a large extent, this computational bottleneck is because existing approaches for computing statistics
typically do not incorporate the relatedness amongst samples.

@@

The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change because of the results of recombination in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
\citep{kelleher2016efficient} used this redundancy
to store genome sequence and the associated genealogical relationships extremely compactly,
as well as to compute several population genetics statistics efficiently.
This paper generalizes the strategy used by \citep{kelleher2016efficient} for statistics computation
to a much broader class of statistics.
